Does It Scale? Symbolic Execution of PHP Web Applications

### 2017-01-23 paper.tex
2 refactored

During the introduction of new semantics for object-oriented programming we
faced a trade-off between accuracy and scalability: Whilst exploring multiple
paths, for method invocations the number of target objects may grow rapidly and
infeasible.

Moreover, based on observations we identify limitations of symbolic execution
for PHP. Dynamic features such as dynamic function invocations and dynamic
include for an accurate execution require precisely resolved include targets
and function names respectively. Since these expressions  can contain symbolic
values at run-time, for many cases this restricts a symbolic execution engine
from further execution as callable code sections become inaccessible.
Although some effort can be spent to statically resolve dynamic include
expressions \cite{hills2014static}, or to simply provide static counterparts manually, resolution
is often not trivial since expressions can contain symbolic values assumed instead
of concrete program input, including databases, user inputs or configuration
files. We empirically evaluate our symbolic execution engine for a corpus
including large and actively maintained PHP systems.

@@

\section{State Of the Art}\label{sec:stateoftheart}
This section recaps the idea of symbolic
execution, the tools which make use of it, and both conceptual and
language-related limitations.

@@

Execution starts with a tautology path
condition. The control-flow can be split whenever the decision between branches
is ambiguous. This is the case, when conditional expressions, for example for
if-statements, evaluate to a symbolic value rather than an actual result.
The corresponding path conditions (guards for instance) are conjoined with the
previous path condition \cite{King1976}. Additionally, the conjoined path
condition can be checked for satisfiability in order to exclude infeasible
paths.

@@

Figure \ref{fig:code_example} shows two code listings. The snippet of server-side web
application code in Figure \ref{fig:web_application} contains both HTML code
(input form in lines 1 -- 4) as well as PHP code: First, the function
\texttt{make\_titles} defined in lines 7 – 13 prints the string \texttt{\$arg} passed as an argument between zero and an arbitrary number of
times. Second, the function is indirectly called using the built-in mechanism
\texttt{call\_user\_func} in line 15 with the string \texttt{Headline} and
\texttt{4} as arguments.  Finally, the input from the form in lines 1 – 4 is
used: The variable \texttt{\$greeting} consists of three concatenated strings,
\texttt{Hello}, the value of input field \texttt{name}, and an exclamation mark.
For the greeting there are two possible client page output
variants: If the input field \texttt{name} is empty at run-time, line 19 is
executed and \texttt{No name entered!} is printed, or, if a name was entered,
line 21 is executed and the greeting constructed in line 17 is printed.

The following symbolic execution semantics for PHP features based on
previous work \cite{Nguyen:2014:BCG:2635868.2635928} illustrate the necessary
modifications and extensions made to handle the level of ambiguity that comes with handling symbolic values. In addition to normal output, the
output approximation contains preprocessor directives to express
repeatable parts as well as output variants. All approximated client page
output corresponding to Figure \ref{fig:web_application} consisting of two different variants
is illustrated in Figure \ref{fig:output_example}.

First, for loops in the previous execution semantics
\cite{Nguyen:2014:BCG:2635868.2635928} only one iteration is executed instead of
an arbitrary number of iterations. Since a loop condition can contain symbolic
values, it might not be feasible to determine whether, and if so, when a loop
execution terminates. Hence, any output constructed during the single loop
iteration is highlighted (\texttt{repeat \ldots endrepeat}) to possibly be
repeated an arbitrary number of times. The output in line 7 is marked to be
repeated an arbitrary number of times, as the number of loop iterations cannot
be determined statically.

Second, a similar procedure applies to recursive
functions. Once called, a recursive function is  also only executed to recursion
depth one per call, and any subsequent invocation of that function returns a
symbolic value. This, again, is due to the difficulty to determine the exact
recursion depth similar to the number of loop iterations.

Third, any form of user interaction providing input data as well as any form of input referred
to from the deployment context of the web application, i.e. configuration files
or databases, is assumed to be symbolic. The context variable
\texttt{\$\_POST} represents input submitted to the server and, hence, is
symbolic as every program input is per definition substituted by symbolic values. Finally, the
output in lines 11 and 13 respectively are marked as alternative variants
(\texttt{if \ldots else \ldots endif}) as symbolic execution of lines 18 – 22
in Figure \ref{fig:web_application} will explore both different branches.

\subsection{Symbolic-Execution-based Tools For PHP}\label{sec:existing_tools}
All tools proposed so far leverage a representation of all possible HTML client
page output. Since any single client page output variant can be analyzed this
enables tool support addressing the web application as a whole in spite of its
staged nature, where parts of the web application are
generated dynamically.
Based on the approximated output representation, subsequent analyses can be
conducted:
\begin{itemize}
	\item \emph{PhpSync}: Using all variants of  HTML client page output, every
	single one can be statically checked for markup validity, i.e., if the web page
	conforms to syntactical specifications for HTML and other client-side
	languages. After the tool detects validation errors, auto-fixes
	can be provided. Otherwise, it traces validation errors back to
	source code responsible for the defect, which then can be refactored manually
	\cite{Nguyen:2011:AFH:2190078.2190142}.

	\item \emph{WebSlice}: Program slices enable to extract and understand the
	impact of changes in an application. To consider all of a dynamic web
	application for program slices, client page output needs to be taken into
	account. WebSlice combines PHP data-flow information with an output
	approximation to enable program slices across different languages
	\cite{Nguyen:2015:CPS:2786805.2786872}.

	\item \emph{Varis}: Editor services such as “jump to declaration” are
	nontrivial for dynamic web applications due to their staged nature. Varis
	provides editor services across stages for client-side code: Starting with an
	output approximation a call-graph with conditional edges is constructed, which
	allows navigation in client-site code (HTML, JS and CSS) although it is
	embedded in server-side code
	\cite{Nguyen:2015:VIS:2819009.2819140,Nguyen:2014:BCG:2635868.2635928}.

@@

\subsection{Assumptions And Limitations}\label{sec:limitations}
In addition to the assumptions introduced by symbolic execution, such as
symbolic inputs and one-time loop execution, the symbolic execution
engine described in previous work \cite{Nguyen:2014:BCG:2635868.2635928} does
not support advanced language features, including object-oriented programming. Moreover, the range of
standard library functions supported are mostly string operations, yet PHP
provides a large collection of functions for array operations.
All three analyses described in Section \ref{sec:existing_tools} though have
only been applied to systems that are small and not maintained any more. Despite the
functional benefits these tools provide to developers, their practicality for
large and modern systems with respect to advanced language support has not been
evaluated.

>>>>

During the introduction of new semantics for object-oriented programming we
faced a trade-off between accuracy and scalability: Whilst exploring multiple
paths, for method invocations the number of target objects may grow rapidly and
infeasible. Moreover, based on observations we identify limitations of symbolic execution
for PHP. Dynamic features such as dynamic function invocations and dynamic
include for an accurate execution require precisely resolved include targets
and function names respectively. Since these expressions  can contain symbolic
values at run-time, for many cases this restricts a symbolic execution engine
from further execution as callable code sections become inaccessible.
Although some effort can be spent to statically resolve dynamic include
expressions \cite{hills2014static}, or to simply provide static counterparts manually, resolution
is often not trivial since expressions can contain symbolic values assumed instead
of concrete program input, including databases, user inputs or configuration
files. We empirically evaluate our symbolic execution engine for a corpus
including large and actively maintained PHP systems.

@@

\section{State Of the Art}\label{sec:stateoftheart}
This Section recaps the idea of symbolic
execution, the tools using output approximation based on it and whose evaluation
we criticize, and current conceptual and language-related limitations.

@@

Symbolic execution commences with a tautology path condition. The control-flow
can be split whenever the decision between branches is ambiguous. This is the
case, when conditional expressions, for example for if-statements, evaluate to
a symbolic value rather than an actual result. The corresponding branch
predicates are conjoined with the previous path condition \cite{King1976}.
Additionally, the conjoined path condition can be checked for satisfiability in
order to exclude infeasible paths.

@@

Figure \ref{fig:code_example} shows two code listings. The snippet of server-side web
application code in Figure \ref{fig:web_application} contains both HTML code
(input form in lines 1 -- 4) as well as PHP code: First, the function
\texttt{make\_titles} defined in lines 7 – 13 prints the string \texttt{\$arg} 
passed as an argument between zero and an arbitrary number of times. Second, the
function is dynamically invocated using the built-in mechanism
\texttt{call\_user\_func} in line 15 with the string \texttt{Headline} and \texttt{4} as arguments.  
Finally, the input from the form in lines 1 – 4 is used: The variable \texttt{\$greeting} consists of three concatenated strings,
\texttt{Hello}, the value of input field \texttt{name}, and an exclamation mark.
For the greeting there are two possible client page output
variants: If the input field \texttt{name} is empty at run-time, line 19 is
executed and \texttt{No name entered!} is printed. Otherwise, if a non-empty
name was entered, line 21 is executed and the greeting constructed in line 17 is printed.

The following symbolic execution semantics for PHP features based on
previous work \cite{Nguyen:2014:BCG:2635868.2635928} illustrate the necessary
modifications and extensions made to handle the level of ambiguity that comes with handling symbolic values. 
In addition to normal output, the output approximation illustrated in
Figure \ref{fig:output_example} contains preprocessor directives to express
repeatable parts as well as output variants. All approximated client page
output corresponding to Figure \ref{fig:web_application} consisting of two
different variants is illustrated in Figure \ref{fig:output_example}.

First, for loops in the previous execution semantics
\cite{Nguyen:2014:BCG:2635868.2635928} only one iteration is executed instead of
an arbitrary number of iterations. Since a loop condition can contain symbolic
values, it might not be feasible to determine whether, and if so, when a loop
execution terminates. Hence, any output assembled during the single loop
iteration is highlighted (\texttt{repeat \ldots endrepeat}) to possibly be
repeated an arbitrary number of times. The output in line 7 is marked to be
repeated an arbitrary number of times, as the number of loop iterations cannot
be determined statically.

Second, a similar procedure applies to recursive
functions. Once invocated, a recursive function is also only executed to
recursion depth one per call, and any subsequent invocation of that function returns a
symbolic value. Again, this is due to the difficulty to determine the exact
recursion depth similar to the number of loop iterations.

Third, any form of user interaction providing concrete input data as well as any
form of input referred to from the deployment context of the web application, i.e. configuration files
or databases, is assumed to be symbolic. The PHP context variable
\texttt{\$\_POST} represents input submitted to the server and, hence, is
symbolic as every program input is per definition substituted by assumed
symbolic values. Finally, the output in lines 11 and 13 respectively are marked as alternative variants
(\texttt{if \ldots else \ldots endif}) as symbolic execution of lines 18 – 22
in Figure \ref{fig:web_application} will explore both different branches.

\subsection{Symbolic-Execution-based Tools For PHP}\label{sec:existing_tools}
All tools proposed so far and presented in the following leverage symbolically
approximated client page output. Since any single client page output variant can
be analyzed this enables tool support addressing the web application as a whole
in spite of its staged nature, where parts of the web application are
generated dynamically.
Based on the approximated output representation, subsequent developer tools and
analyses have been proposed:
\begin{itemize}
	\item \emph{PhpSync}: Using all variants of HTML client page output, every
	single one can be statically checked for markup validity, i.e., if the web page
	conforms to syntactical specifications for HTML and other client-side
	languages. After the tool detects validation errors, if possible, auto-fixes
	are provided. Otherwise, it traces validation errors back to
	source code responsible for the defect, which then can be refactored manually
	\cite{Nguyen:2011:AFH:2190078.2190142}.

	\item \emph{WebSlice}: Program slices enable to extract and understand the
	impact of changes in an application. To consider all of a dynamic web
	application for program slices, client page output needs to be taken into
	account. \sf{WebSlice} combines PHP data-flow information with an output
	approximation to enable program slices across different languages
	\cite{Nguyen:2015:CPS:2786805.2786872}.

	\item \emph{Varis}: Editor services such as “jump to declaration” are
	not trivial for dynamic web applications due to their staged nature. \sf{Varis}
	provides editor services across stages for client-side code: Starting with an
	output approximation a call-graph with conditional edges is constructed, which
	allows navigation in client-site code (HTML, JS and CSS) although it is
	embedded in server-side code
	\cite{Nguyen:2015:VIS:2819009.2819140,Nguyen:2014:BCG:2635868.2635928}.

@@

\subsection{Assumptions And Limitations}\label{sec:limitations}
In addition to the assumptions introduced by symbolic execution, such as
symbolic inputs and one-time loop execution, the symbolic execution
engine described in previous work \cite{Nguyen:2014:BCG:2635868.2635928} does
not yet support advanced language features, including object-oriented
programming.
Moreover, the range of standard library functions supported are mostly string operations, yet PHP
provides a large collection of functions for array operations.
All three tools described in Section \ref{sec:existing_tools} have
only been applied to systems that are small and not actively maintained.
Despite the functional benefits that these tools can provide to developers,
their practicality for large and modern systems with respect to advanced language support has not been
assessed.
